Improving Haskell Theory Exploration

### 2015-11-30 TransferReport/background.tex
Added Core translations

\begin{figure}
  \begin{equation*}
    \begin{split}
      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
                         |\ & \texttt{Lit}\ literal                                  \\
                         |\ & \texttt{App}\ expr\ expr                               \\
                         |\ & \texttt{Lam}\ \mathcal{L}\ expr                        \\
                         |\ & \texttt{Let}\ bind\ expr                               \\
                         |\ & \texttt{Case}\ expr\ \mathcal{L}\ \left[ alt \right]   \\
                         |\ & \texttt{Type}                                          \\
      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
                         |\ & \texttt{Global}\ \mathcal{G}                           \\
      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
      alt\     \rightarrow\ & ( altcon,\ [\mathcal{L}],\ expr )                      \\
      altcon\  \rightarrow\ & \texttt{DataAlt}\ \mathcal{G}                          \\
                         |\ & \texttt{LitAlt}\ literal                               \\
                         |\ & \texttt{Default}                                       \\
      bind\    \rightarrow\ & \texttt{NonRec}\ \mathcal{L}\ expr                     \\
                         |\ & \texttt{Rec}\ [ ( \mathcal{L},\ expr ) ]
    \end{split}
  \end{equation*}
  Where:
  \begin{tabular}[t]{l @{ $=$ } l}
    $\mathcal{S}$ & string literals    \\
    $\mathcal{N}$ & numeric literals   \\
    $\mathcal{L}$ & local identifiers  \\
    $\mathcal{G}$ & global identifiers
  \end{tabular}

@@

The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to ignore type hints (explicit casts are removed, whilst particular types/kinds/coercions are all represented by \texttt{Type}) and to use a custom representation of names. We also omit several other forms of literal (machine words of various sizes, individual characters, etc.) for brevity, as their treatment is similar to those of strings and numerals.

@@

@@

\newcommand{\blank}{\cdot}

@@

\subsection{Clustering and Feature Extraction}
\label{sec:featureextraction}

Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources. The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs. We describe this approach in more details in \S \ref{sec:relevance}. Premise selection is a practical tool, used in production systems like the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting. The reason is that theory exploration has no distinct \emph{goal} to compare expressions against. Instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}. A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way. See \S \ref{sec:clustering} for more information.

@@

\iffalse

@@

\subsection{Clustering}
\label{sec:clustering}
% TODO
\fi

>>>>

\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}

\begin{figure}
  \begin{equation*}
    \begin{split}
      expr\    \rightarrow\ & \CVar\ id                                       \\
                         |\ & \CLit\ literal                                  \\
                         |\ & \CApp\ expr\ expr                               \\
                         |\ & \CLam\ \mathcal{L}\ expr                        \\
                         |\ & \CLet\ bind\ expr                               \\
                         |\ & \CCase\ expr\ \mathcal{L}\ \left[ alt \right]   \\
                         |\ & \CType                                          \\
      id\      \rightarrow\ & \CLocal\ \mathcal{L}                            \\
                         |\ & \CGlobal\ \mathcal{G}                           \\
      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}                           \\
                         |\ & \CLitStr\ \mathcal{S}                           \\
      alt\     \rightarrow\ & ( altcon,\ [\mathcal{L}],\ expr )               \\
      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{G}                          \\
                         |\ & \CLitAlt\ literal                               \\
                         |\ & \CDefault                                       \\
      bind\    \rightarrow\ & \CNonRec\ \mathcal{L}\ expr                     \\
                         |\ & \CRec\ [ ( \mathcal{L},\ expr ) ]
    \end{split}
  \end{equation*}
  Where:
  \begin{tabular}[t]{l @{ $=$ } l}
    $\mathcal{S}$ & string literals    \\
    $\mathcal{N}$ & numeric literals   \\
    $\mathcal{L}$ & local identifiers  \\
    $\mathcal{G}$ & global identifiers
  \end{tabular}

@@

The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions). We also omit several other forms of literal (machine words of various sizes, individual characters, etc.) for brevity, as their treatment is similar to those of strings and numerals. We denote names and literals using quoted strings, e.g. \texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{LitStr "baz"} and \texttt{LitNum "42"}.

\begin{figure}
  \begin{haskell}
    data Nat = Z
             | S Nat

    plus :: Nat -> Nat -> Nat
    plus    Z  y = y
    plus (S x) y = S (plus x y)

    mult :: Nat -> Nat -> Nat
    mult    Z  y = Z
    mult (S x) y = plus y (mult x y)

    factorial :: Nat -> Nat
    factorial    Z  = S Z
    factorial (S n) = mult (S n) (factorial n)

    odd :: Nat -> Bool
    odd    Z  = False
    odd (S n) = even n

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{haskell}
  \caption{A Haskell datatype and functions for Peano arithmetic; in particular, the \hs{even} and \hs{odd} functions are mutually recursive. The \hs{Bool} type is Haskell's built in representation of Booleans, which can be regarded as \hs{data Bool = True \| False}.}
  \label{fig:haskellexample}
\end{figure}

\begin{figure}
  \begin{subfigure}[plus]{\textwidth}
    \begin{small}
      \underline{\texttt{plus}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       [(DataAlt "Z", [],    Var (Local "y")),
                        (DataAlt "S", ["x"], App (Var (Global "S"))
                                                 (App (App (Var (Global "plus"))
                                                           (Var (Local  "x")))
                                                      (Var (Local "y"))))]))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[mult]{\textwidth}
    \begin{small}
      \underline{\texttt{mult}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       [(DataAlt "Z", [],    Var (Global "Z")),
                        (DataAlt "S", ["x"], App (App (Var (Global "plus"))
                                                      (Var (Local  "y")))
                                                 (App (App (Var (Global "mult"))
                                                           (Var (Local  "x")))
                                                      (Var (Local  "y"))))]))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[factorial]{\textwidth}
    \begin{small}
      \underline{\texttt{factorial}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              [(DataAlt "Z", [],    App (Var (Global "S"))
                                        (Var (Global "Z"))),
               (DataAlt "S", ["n"], App (App (Var (Global "mult"))
                                             (App (Var (Global "S"))
                                                  (Var (Local  "n"))))
                                        (App (Var (Global "factorial"))
                                             (Var (Local  "n"))))])
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[odd]{\textwidth}
    \begin{small}
      \underline{\texttt{odd}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              [(DataAlt "Z", [],    Var (Global "False")),
               (DataAlt "S", ["n"], App (Var (Global "even"))
                                        (Var (Local  "n")))])
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[even]{\textwidth}
    \begin{small}
      \underline{\texttt{even}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              [(DataAlt "Z", [],    Var (Global "True")),
               (DataAlt "S", ["n"], App (Var (Global "odd"))
                                        (Var (Local  "n")))])
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \caption{Translations of functions in figure \ref{fig:haskellexample} into the Core syntax of figure \ref{fig:coresyntax}. Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions. Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
  \label{fig:coreexample}
\end{figure}

Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals. The translation to our Core syntax is routine, and shown in figure \ref{fig:coreexample}. Although the Core is more verbose, we can see similar structure across the definitions (such as \texttt{(App (App (Var (Global $\blank$)) (Var (Local "x"))) (Var (Local "y")))}) which arise from similarities in the Haskell definitions (in this case, \hs{(plus x y)} and \hs{(mult x y)}). This is clearest in the definitions of \hs{odd} and \hs{even}, which are identical except for the global variables.

Note that we exclude representations for type-level entities, including datatype definitions like that of \hs{Nat}. GHC can represent these, but in this work we only consider reducible expressions (i.e. value-level bindings of the form \hs{f a b ... = ...}).

@@

\iffalse TODO: Re-use these examples later on, e.g. for QuickSpec exploration \fi

@@

@@

\subsection{Feature Extraction}
\label{sec:featureextraction}

\iffalse TODO: include definitions \fi
\iffalse TODO: make more relevant to our context \fi

Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources. The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs. We describe this approach in more details in \S \ref{sec:relevance}. Premise selection is a practical tool which has seen use in practical software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting. The reason is that theory exploration has no distinct \emph{goal} \iffalse TODO: implies that premise selection has a goal; rephrase: premise selection is given a concrete formula (called the ``goal'') to compare terms against \fi to compare expressions against. Instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}. A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way. See \S \ref{sec:clustering} for more information.

@@

@@

\subsection{Clustering}
\label{sec:clustering}
% TODO

Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric. There are many variations on this theme, but in our case we will fix the number of clusters at \iffalse TODO: use ceiling brackets \fi $k = \sqrt{n}$ the data points will be $d$-dimensional feature vectors of rational numbers, the similarity metric is euclidean distance, denoted $d$. and clustering will be performed by the $k-means$ algorithm. We choose this setup as it is well studied and supported by off-the-shelf tools, and is used in the ML4PG and ACL2(ml) systems which served as inspiration for this work.

We will use $S$ to denote our set of $n$ data points and $k$ as the number of clusters.

\subsubsection{k-Means}

The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster's elements, which we will denote as $m_1$ to $m_k$. Since k-means is iterative, we will use superscript notation to denote time steps, such as $m_i^t$ for the value of $m_i$ at step $t$.

Before k-means starts, we must choose \emph{seed} values for $m_1^0$ to $m_k^0$. Many methods have been proposed for choosing these values, e.g. see CITE. For simplicity, we will assume they are chosen at random from our data set. Subsequent values, i.e. $m_i^t$ for $t > 0$, are simply the mean values of the corresponding cluster $C_i^t$:

TODO: We need to handle the vectors element-wise, but we've already used subscript notation!

\begin{equation}
  m_i^t =
\end{equation}

At time step $t$, the elements of each cluster $C_i^t$ are the data points closest to $m_i^{t-1}$ from the previous time step:

\begin{equation}
  C_i^t = \{ p \in S \mid i = \argmin\limits_j d(p, m_j^{t-1}) \}
\end{equation}

The mean values o determine which cluster a point $p$ belongs to, we
The as the name suggests, relies on the mean value of each cluster to determine  represents each cluster by uses a \emph{representative} t step $t$, the with cluster membership  We begin at iteration 0 by \emph{seeding} the algorithm with , each cluster is represented by the mean value of its elements:

\begin{equation}
  C_k =
\end{equation}
